Depression and suicidality as evolved credible signals of need in social conflicts

### 2021-06-09 Paper.rmd
* Edits to tie sections together * Added section on perceived mental illness * Two new effects plots on the effect of mental illness perceptions on belief and action

---
title: "Depression and suicidality as credible signals of need in social conflicts"
author:
- name: Michael R. Gaffney
  affiliation: Department of Anthropology, Washington State University
- name: Kai Adams
  affiliation: UC Berkeley Haas School of Business
- name: Kristen L. Syme
  affiliation: Department of Experimental and Applied Psychology, Vrije Universiteit Amsterdam
- name: Edward H. Hagen
  affiliation: Department of Anthropology, Washington State University
  email: edhagen@wsu.edu
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: true
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
    css: style.css
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

Mental health professionals generally view major depression and suicidality as pathological responses to stress that elicit aversive responses from others. An alternative hypothesis grounded in evolutionary theory contends that depression and suicidality are honest signals of need in response to adversity that can increase support from reluctant others when there are conflicts of interest. To test this hypothesis, we examined responses to emotional signals in an experimental vignette study involving claims of substantial need in the presence of conflicts of interest and private information about the signaler's true level of need. In a sample of 1,240 participants recruited from Amazon Mechanical Turk, costlier signals like depression and suicidality resulted in greater perceptions of need, reduced perceptions of manipulativeness, and increased likelihood of support compared to simple verbal requests and crying without further symptoms. Additionally, as predicted, the effect of signaling on likelihood of support was largely mediated by the effect of signaling on participants' belief that the signaler was genuinely in need. Our results support the hypothesis that depression and suicidality, apparent human universals, are credible signals of need.

@@

In a classic study, @Coyne1976 found that depression alienates others, a result that was subsequently confirmed in many other studies [@segrin1992; @segrin2000]. In interactions with spouses, depressed individuals express anger and aggression, and make frequent demands for help [@segrin1994]. Depressed individuals also self-disclose personally relevant negative issues at inappropriate times and they view such topics as more appropriate for discussion than do the non-depressed [@segrin2000]. Such self-disclosures have been shown to be a key ingredient in the rejection of depressed persons by others, and "may appropriately be understood as an attempt to elicit social support from targets" [@segrin1994, p. 657]. Excessive reassurance seeking -- repeatedly requesting reassurance that one is lovable and worthy despite previous attempts by others to provide such reassurance -- is another factor implicated in the rejection of the depressed by others [@joiner1999; @JoinerMetalsky1995; @StarrDavila2008].

Negative social responses to depression are widely interpreted as evidence of impaired social functioning in the depressed [@EvraireDozois2011; @GotlibLee1989; @HirschfeldEtAl2000], including by evolutionary theorists of depression [@nettle2004; @allen2006; @nesse2006; @myers2016]. This interpretation is reinforced by the equally widespread view that depressed individuals have impaired, negative perceptions of themselves and their environments [@beck1963; @joiner2007; @nolen1991]. Together with the costs of depression, such as suicide, these facts are the basis of the mainstream claim that depression is a psychopathology.

In contrast to these views, we will argue that the depressed have suffered genuinely severe adversities and are therefore genuinely in need, but have conflicts with their social partners. In these circumstances, aversive but credible signals of need can evolve.

## Depression is probably caused by genuine adversity

@@

Contrary to the view that the depressed have a distorted perception of their environment, there is strong evidence that most cases of depression are caused by genuinely severe negative life events, such as physical assault and death of a loved one [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008; @devries2013]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. Longitudinal studies indicate that depression onsets soon after a negative event [@lewinsohn1988; @KendlerEtAl1999; @rich2005; @han2019; @sen2010] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after, @blekesaune2008; @metsa2013; @rosenstrom2017]. Although the relationship between negative events and depression is likely bidirectional [i.e., depression probably also causes adversity, @wichers2012], negative events predict depression even when considering only events outside of one’s control, indicating that the connection is unlikely to be driven solely by individuals who are already depressed selecting into situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Furthermore, twin studies have shown that one's history of negative events remains a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999]. For these and other reasons, most cases of depression are probably functional instances of psychological pain, i.e., the severe end of a spectrum of low mood, sadness, and grief, and not mental dysfunctions [@Hagen2003; @horwitz_loss_2007; @andrews2010; @SymeHagen2021; @dowrick2013; @frances2013].

@@

There is equally strong evidence that depression is closely associated with anger and conflict [@cassiello2016]. Of the types of negative life events associated with depression, those which involve conflict tend to be more strongly associated with depression [@Hammen2005; @Mazure1998]. Marital problems, bullying, and abusive relationships are all common risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with assault (both sexual and non-sexual), in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999]. This holds true even in a small-scale, non-Western society: among the Tsimane, Amazonian horticulturalists, depression is also associated with conflict, particularly conflict involving non-kin [@StieglitzEtAl2015a]. See @SymeHagen2021 for a review of the association of depression with anger and conflict.

Other notable risk factors, like loss of a loved one or severe or prolonged illness, might seem less related to conflict. In these situations, however, the fitness costs that stem from reduced access to resources could be mitigated with help from others [@SugiyamaSugiyama2003] who might not be able, or want, to provide more investment than they already are. Indeed, there is increasing evidence that loss of a loved one is often followed by increased family conflict [see @SymeHagen2021 for a brief review]. Problems whose solutions require substantial investment or other changes on the part of social partners will therefore often involve social conflict even if they did not start that way [@Hagen2003]. When there are severe conflicts of interest, individuals have incentives to deceive, so "cheap" signals of need, such as sad expressions and crying, might not be believed.

@@

@Hagen2003 proposed that physical aggression and core depression symptoms like loss of interest in virtually all activities were complementary strategies to resolve conflicts in interdependent relationships. @SellEtAl2009 found that physically formidable individuals were more prone to anger, prevailed more in conflicts of interest, and considered themselves entitled to better treatment. Physically or socially weaker individuals, though, are not without options to resolve conflicts in their favor. An individual with monopoly power over the benefits she contributes to a critical cooperative endeavor can withhold those benefits, or put them at risk, until her partners change their behaviors in ways that benefit her. Depression is characterized, in part by a profound loss of interest in virtually all activities [@dsm5], a symptom that is probably responsible for a sharp reduction in productivity [for review, see @SymeHagen2021]. Depression might therefore be a bargaining strategy for relatively powerless individuals in the wake of adversity and social conflict [@Hagen2003; @SymeHagen2021; see also @Watson:2002].

@@

Suicidal ideation is one of nine diagnostic criteria for a major depressive episode (MDE) [@dsm5], it is associated with depression across cultures [@haroz2017], and depression is a major risk factor for suicidal behavior [@hawton2013].

Anthropology has long viewed suicidality as largely the result of social problems. Early in the field's history, anthropologists reported on suicide attempts and deaths in the small-scale societies in which they conducted their fieldwork. Suicide, they found, was commonly a form of protest, revenge, and/or appeal [@firth1936; @Firth:1961; @Malinowski:1932; @niehaus2012]. Some ethnographers emphasized suicide as a form of anger or social pressure [@Hezel:1987; @Giddens:1964], whereas others emphasized the powerlessness of suicide victims [@Counts:1980].

Common to almost all theoretical and empirical investigations of suicide is a focus on *completed* suicides, i.e., suicide deaths. The vast majority of suicidal behavior, however, does not result in death. In young adult women in the US, for example, there are hundreds of attempts for every death (see Figure \@ref(fig:cdc)). @SymeEtAl2016 therefore argued that the theoretical focus should be on suicide ideation and suicide attempts.

@@

In the bargaining framework, suicidality is conceptualized as putting all future contributions to cooperative endeavors with social partners at risk with some low but non-zero probability, credibly signaling low valuation of current circumstances [see @Rosenthal1993 for a game theory model of suicide attempts as costly signals]. On this view, most suicides deaths, especially in young, physically healthy individuals, would therefore be the inevitable consequence that some individuals lose this gamble. These and similar ideas [@andrews2006; @HagenEtAl2008; @nock2008; @Stengel1956] suggest that depression symptoms that put one's value to others at risk, such as loss of interest and suicidality, are credible signals of need.

<!-- Need to integrate this -->
Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are all being commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012]. Despite this potential for stigmatization, increased social support and changes to important relationships have been reported to follow suicide attempts with some indication that these effects may hold long term [@Stengel1956]. For example, a study of 100 women who survived suicide attempts found that individuals gained identifiable benefits through the attempt in 75 cases, with 41 individuals benefiting from reconciliations with others [@Lukianowicz1971]. Unlike many Western countries, where suicide is often viewed as pathological [@Hidaka2012], members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness [@Shostak1981], with attempts having been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers [@GutierrezdePinedaMuirden1948; @Hilger1957; @Karsten1935; @Tessmann1930; @Wilson1960]. Although findings that individuals view suicide attempts as cries for help is not necessarily evidence for their ability to lead to beneficial responses, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor, despite suicide rarely if ever being the main topic [@SymeEtAl2016].

Some evolutionary theorists have questioned the bargaining model based on evidence that depression is aversive to social partners [@nettle2004; @allen2006; @nesse2006; @myers2016]. These critics overlooked the central role of *conflict* in the model. Anger is aversive to targeted social partners, yet it is probably an adaptation that exploits advantages in physical or social formidability to force concessions from them [@SellEtAl2009]. Aversive responses to depressive bargaining are likewise expected throughout the process, with potential helpers signaling the costliness of providing support in turn. Importantly, as with anger, aversive responses to depression encourage concessions by interdependent social partners [@SymeHagen2021].

@@

The prevailing view is that depression causes social rejection. The aim of this study was to test the hypothesis that when there are conflicts of interest, depressive and suicidal behaviors increase belief that victims of adversity are telling the truth and consequently increase willingness to help them.

<!-- Beneficial responses to depressed individuals have also been reported, such as increased caretaking [@HokansonEtAl1986], more offers of advice and support [@stephens1987], and reduced aggression within families [@hops1987; @dadds1992; @SheeberEtAl2001].  -->

@@

At baseline (T1), `r mentallyIll`% of participants thought the victim was mentally ill, which decreased slightly at T2 in the *Verbal request* and *Crying* conditions (`r miT2$percent[miT2$signal=='Crying']`%), and then increased with signal cost to `r miT2$percent[miT2$signal=='Mild depression']`% in *Mild depression*, `r miT2$percent[miT2$signal=='Depression']`% in *Depression*, and `r miT2$percent[miT2$signal=='Suicide attempt']`% in *Suicide attempt*, with some variation by vignette. See Figures \@ref(fig:mentallyillT1) and \@ref(fig:mentallyillT2).

@@

As predicted, in vignettes involving conflicts of interest and private information about the need for help, costly signals of need increased participants' belief in the victim and their likelihood of helping her, with the increase in perceived need and likelihood of helping increasing monotonically with signal cost. As predicted, the increase in likelihood of helping was largely mediated by the increase in belief in the victim. In an exploratory analysis, costlier signals also decreased perceptions that the victim was manipulative.

These effects were largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters, and smaller in the "basketball coach" and the "thwarted marriage" vignettes. The small effect in the "basketball coach" vignette might have been due to the lack of relatedness between participants and the victim (the star player), or how participants weighted the costs of suspending the coach vs. punishing a potentially innocent person. The US vignettes also had different degrees of evidence *against* the victim beyond just denial by the accused, ranging from strong evidence in the basketball coach vignette (a negative police report) to moderate evidence in the romantic partner vignette (no physical injuries) to weak evidence in the brother-in-law vignette (nothing beyond denial by the brother-in-law), raising the possibility that credible signals are more effective when negative evidence is lacking (Dylan Tweed, personal communication).

The small signal effect in the "thwarted marriage" vignette, which involved the Indian sample, could indicate that our results do not generalize across cultures, undermining our adaptationist hypothesis. It could also reflect our poor understanding of contemporary Indian culture regarding dowry (the effect was larger in older participants). Baseline belief in the older daughter, and likelihood of helping her, was relatively high at baseline (`r signif(100*mean(d$T1Belief[d$vignette=='Thwarted marriage']), 2)`%) compared to victims in the other vignettes. Private information and conflict therefore probably played a smaller role and thus costly signals were less necessary. We observed a similar pattern in our pilot study, in which baseline belief in the victim's need was high, and costly signals had smaller effects than they did in the current study. Additionally, supporting the older daughter came at the cost of one's younger daughter, which may also help explain the relatively small signal effects. A final consideration is that data from the Indian sample appeared to be of lower quality, limiting our confidence in any of these interpretations (see limitations for more information).

In the "brother-in-law" and "romantic partner" scenarios, *Crying* had little effect on the magnitude of pro-victim responses relative to *Verbal request*, suggesting it was not costly enough to serve as a reliable signal in times of substantial conflicts of interests. In contrast, both *Depression* conditions increased support, albeit to similar degrees. One potential reason for this is the increase in symptoms from mild depression to depression was not costly enough to result in a large effect, as the symptoms differed little in severity (e.g., grades dropping from As to Bs in *Mild depression* vs. Cs in *Depression*). Such small changes may be less impactful in vignettes than in real-life, where the effects of signaling may increase in severity as they persist over a lengthy period of time.

@@

@@

Depression is costly and sometimes leads death by suicide. Our results suggest that these costs, which mainstream theories take as evidence of brain dysfunction [@SymeHagen2021], instead function to help victims of adversity elicit support when their true level of need is private information and they have conflicts with social partners. These findings align with real-world evidence that depression and suicide simultaneously elicit positive and negative responses from social partners [@SymeHagen2021]. If the bargaining model is correct, depression and suicidality are important social problems, not medical ones [@syme2020].

@@

In the MTurk study, participants read a vignette that put them in situations in which an imagined sister was requesting a loan of \$50,000 from the participant that was originally saved for the participant's imagined daughter's college education. The sister claims the money is needed for medical treatment for her own child, but she has incentives to deceive. The manipulations were signals very similar to those used in this study, and the outcomes were belief that the sister really needed the money and intention to loan her the money. See XX for the full write-up. See <https://osf.io/3rg8b> for preregistration of the pilot study. See <http://doi.org/10.5281/zenodo.4637883> for pilot data.

@@

```{r mentallyillT2, fig.width=8, fig.cap="Proportion of participants in each condition who thought that the victim was mentally ill at T2, controlling for perceptions of mental illness at T1. Means estimated with a generalized linear regression model with the quasibinomial family."}
effectsPlots$m21plot + ggtitle('Time 2')
```

>>>>

---
title: "Depression and suicidality as evolved credible signals of need in social conflicts"
author:
- name: Michael R. Gaffney
  affiliation: Department of Anthropology, Washington State University
- name: Kai Adams
  affiliation: UC Berkeley Haas School of Business
- name: Kristen L. Syme
  affiliation: Department of Experimental and Applied Psychology, Vrije Universiteit Amsterdam
- name: Edward H. Hagen
  affiliation: Department of Anthropology, Washington State University
  email: edhagen@wsu.edu
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: true
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
    css: style.css
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

Mental health professionals generally view major depression and suicidality as pathological responses to stress that elicit aversive responses from others. An alternative hypothesis grounded in evolutionary theory contends that depression and suicidality are honest signals of need in response to adversity that can increase support from reluctant others when there are conflicts of interest. To test this hypothesis, we examined responses to emotional signals in a preregistered experimental vignette study involving claims of substantial need in the presence of conflicts of interest and private information about the signaler's true level of need. In a sample of 1,240 participants recruited from Amazon Mechanical Turk, costlier signals like depression and suicidality resulted in greater perceptions of need, reduced perceptions of manipulativeness, and increased likelihood of support compared to simple verbal requests and crying without further symptoms. Additionally, as predicted, the effect of signaling on likelihood of support was largely mediated by the effect of signaling on participants' belief that the signaler was genuinely in need. Our results support the hypothesis that depression and suicidality, apparent human universals, are credible signals of need that elicit more support than verbal requests, sad expressions, and crying when there are conflicts of interest.

@@

In a classic study, @Coyne1976 found that depression alienates social partners, a result that was subsequently confirmed in numerous studies [@segrin1992; @segrin2000]. In interactions with spouses and others, depressed individuals express anger and aggression, make frequent demands for help, self-disclose personally relevant negative issues at inappropriate times, and view such topics as more appropriate for discussion than do the non-depressed [@segrin1994; @segrin2000]. Such self-disclosures have been shown to be a key ingredient in the rejection of depressed persons by others, and "may appropriately be understood as an attempt to elicit social support from targets" [@segrin1994, p. 657]. Excessive reassurance seeking -- repeatedly requesting reassurance that one is lovable and worthy despite previous attempts by others to provide such reassurance -- is another factor implicated in the rejection of the depressed by others [@joiner1999; @JoinerMetalsky1995; @StarrDavila2008].

Negative social responses to depression are widely interpreted as evidence of impaired social functioning in the depressed [@EvraireDozois2011; @GotlibLee1989; @HirschfeldEtAl2000; @hames2013; @gadassi2015; @kupferberg2016; @weightman2019]. This interpretation is reinforced by the equally widespread view that depressed individuals have impaired, negative perceptions of themselves and their environments [@beck1963; @joiner2007; @nolen1991]. Together with the costs of depression, such as profound loss of interest in virtually activities and suicide, these facts are the basis of the mainstream claim that depression is a psychopathology.

In contrast to these views, we will argue that the depressed have suffered genuinely severe adversities and are therefore genuinely in need, but often have conflicts with their social partners. In these circumstances, costly and putatively dysfunctional depressive behaviors can instead be understood as aversive but credible and adaptive signals of need that elicit more support than verbal requests, sad expressions, and crying when there are conflicts of interest.

## Depression is caused by genuine adversity

@@

Contrary to the view that the depressed have a distorted perception of their environment, there is strong evidence that most cases of depression are caused by genuinely severe negative life events, such as physical assault and death of a loved one [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008; @devries2013]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. Longitudinal studies indicate that depression onsets soon after a negative event [@lewinsohn1988; @KendlerEtAl1999; @rich2005; @han2019; @sen2010] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after, @blekesaune2008; @metsa2013; @rosenstrom2017]. Although the relationship between negative events and depression is likely bidirectional [i.e., depression probably also causes adversity, @wichers2012], negative events predict depression even when considering only events outside of one’s control, indicating that the connection is unlikely to be driven solely by individuals who are already depressed selecting into situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Furthermore, twin studies have shown that one's history of negative events remains a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999].

For these and other reasons, we and others argue that most cases of depression are probably functional instances of psychological pain, i.e., the severe end of a spectrum of adaptive low mood, sadness, and grief, and not mental dysfunctions [@Hagen2003; @horwitz_loss_2007; @andrews2010; @SymeHagen2021; @dowrick2013; @frances2013].

@@

One might expect that victims of adversity who become depressed would receive positive responses from family, friends, colleagues, and perhaps even strangers. Indeed, beneficial responses to depressed individuals have also been reported, such as increased caretaking [@HokansonEtAl1986], more offers of advice and support [@stephens1987], and reduced aggression within families [@hops1987; @dadds1992; @SheeberEtAl2001]. Why, though, would these positive responses often be accompanied by negative ones?

The missing piece of the puzzle is that depression is closely associated with anger and conflict [@cassiello2016]. Of the adversity risk factors for depression, those that involve conflict tend to be the strongest [@Hammen2005; @Mazure1998]. Marital problems, bullying, and abusive relationships are all common risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with sexual and non-sexual assault, in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999]. This holds true even in a small-scale, non-Western society: among the Tsimane, Amazonian horticulturalists, depression is also associated with conflict, especially conflict involving non-kin [@StieglitzEtAl2015a]. See @SymeHagen2021 for a review of the association of depression with anger and conflict.

Other notable depression risk factors, like loss of a loved one or severe or prolonged illness, might seem less related to conflict. In these situations, however, the fitness costs that stem from reduced access to resources could be mitigated with help from social partners [@SugiyamaSugiyama2003]. However, social partners might not be able, or want, to provide more investment than they already are. Indeed, there is increasing evidence that loss of a loved one is often followed by increased family conflict [see @SymeHagen2021 for a brief review]. Problems whose solutions require substantial investment or other changes on the part of social partners will therefore often involve social conflict even if they did not start that way [@Hagen2003].

Severe adversity, which often requires substantial help from social partners that they might be reluctant to provide, combined with social conflict, often of the he-said-she-said variety that could undermine victims' credibility, render "cheap" signals of need, such as sad expressions and crying, ineffective at increasing support. We argue next that some of the most costly and mysterious symptoms of depression -- profound loss of interest in virtually all activities and suicidality -- are *credible* and adaptive signals of need when there are conflicts of interest between victims of adversity and their social partners.

@@

@Hagen2003 proposed that physical aggression and core depression symptoms like loss of interest in virtually all activities were complementary strategies to resolve conflicts in interdependent relationships. @SellEtAl2009 found that physically formidable individuals were more prone to anger, prevailed more in conflicts of interest, and considered themselves entitled to better treatment. Physically or socially weaker individuals, though, are not without options to resolve conflicts in their favor. An individual with monopoly power over the benefits she contributes to a critical cooperative endeavor can withhold those benefits, or put them at risk, until her partners change their behaviors in ways that benefit her. Depression is characterized, in part by a profound loss of interest in virtually all activities [@dsm5], a symptom that is probably responsible for a sharp reduction in productivity [for review, see @SymeHagen2021]. Depression might therefore be an evolved bargaining strategy for relatively powerless individuals in the wake of adversity and social conflict [@Hagen2003; @SymeHagen2021; see also @Watson:2002].

@@

Suicidal ideation is one of nine diagnostic criteria for a major depressive episode (MDE) [@dsm5] and is associated with depression across cultures [@haroz2017]; depression is a major risk factor for suicidal behavior [@hawton2013]; and suicidality has been invoked to justify grief-related depression symptoms as psychopathological [e.g., @pies2014]. Theoretical models of depression must account for suicidality.

Anthropology has long viewed suicidality as largely the result of social problems. Early in the field's history, anthropologists reported on suicide attempts and deaths in the small-scale societies that serve as models for the types of societies in which humans evolved. Suicide, they found, was commonly a form of protest, revenge, and/or appeal [@firth1936; @Firth:1961; @Malinowski:1932; @niehaus2012]. Some ethnographers emphasized suicide as a form of anger or social pressure [@Hezel:1987; @Giddens:1964], whereas others emphasized the powerlessness of suicide victims [@Counts:1980].

Common to almost all theoretical and empirical investigations of suicide in anthropology and other disciplines is a focus on *completed* suicides, i.e., suicide deaths. The vast majority of suicidal behavior, however, does not result in death. In young adult women in the US, for example, there are hundreds of attempts for every death (see Figure \@ref(fig:cdc)). @SymeEtAl2016 therefore argued that the theoretical focus should be on suicide ideation and suicide attempts.

@@

In the bargaining framework, suicidality is conceptualized as putting all future contributions to cooperative endeavors with social partners at risk with some low but non-zero probability, credibly signaling low valuation of current circumstances. On this view, most suicides deaths, especially in young, physically healthy individuals, would therefore be the inevitable consequence that some individuals lose this gamble. These and similar ideas [@andrews2006; @HagenEtAl2008; @nock2008; @Stengel1956; @Rosenthal1993] suggest that depression symptoms that put one's value to others at risk, such as loss of interest and suicidality, are credible signals of need.

<!-- Need to integrate this -->
As with depression, there are negative social responses to suicidal behavior. Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012].

Despite this potential for stigmatization, increased social support and  beneficial changes to important relationships have been reported to follow suicide attempts with some indication that these effects may hold long term [@Stengel1956]. For example, a study of 100 women who survived suicide attempts found that individuals gained identifiable benefits through the attempt in 75 cases, with 41 individuals benefiting from reconciliations with others [@Lukianowicz1971]. Unlike many Western countries, where suicide is often viewed as pathological [@Hidaka2012], members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness [@Shostak1981], with attempts having been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers [@GutierrezdePinedaMuirden1948; @Hilger1957; @Karsten1935; @Tessmann1930; @Wilson1960]. Although findings that individuals view suicide attempts as cries for help is not necessarily evidence for their ability to lead to beneficial responses, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor [@SymeEtAl2016].

<!-- Some evolutionary theorists have questioned the bargaining model based on evidence that depression is aversive to social partners [@nettle2004; @allen2006; @nesse2006; @myers2016]. These critics overlooked the central role of *conflict* in the model.  -->

Anger is aversive to targeted social partners, yet it is probably an adaptation that exploits advantages in physical or social formidability to force concessions from them [@SellEtAl2009]. Aversive responses to depressive and suicidal bargaining are likewise expected throughout the process, encouraging concessions by interdependent social partners, who in turn signal the costliness of increasing their support [@Hagen2003; @SymeHagen2021].

@@

The prevailing view is that depression involves impaired social abilities that lead to rejection by social partners [@Coyne1976; @joiner1999; @segrin2000; @hames2013; @weightman2019; @gadassi2015]. The aim of this study was to test an alternative hypothesis that when there are conflicts of interest, depressive and suicidal behaviors benefit victims of adversity by increasing belief that they are telling the truth and consequently increasing willingness to help them.

@@

At baseline (T1), `r mentallyIll`% of participants thought the victim was mentally ill, which decreased slightly at T2 in the *Verbal request* and *Crying* conditions (`r miT2$percent[miT2$signal=='Crying']`%), and then increased with signal cost to `r miT2$percent[miT2$signal=='Mild depression']`% in *Mild depression*, `r miT2$percent[miT2$signal=='Depression']`% in *Depression*, and `r miT2$percent[miT2$signal=='Suicide attempt']`% in *Suicide attempt*, with some variation by vignette. See Figures \@ref(fig:mentallyillT1) and \@ref(fig:mentallyillT2). Perceived mental illness at T2 was associated with lower T2 Belief and Action in the US sample, but this effect was mainly evident in the *Verbal request* and *Crying* conditions. See Figures \@ref(fig:mentallyillbelief) and \@ref(fig:mentallyillaction).

@@

As predicted, in vignettes involving conflicts of interest and private information about the need for help, costly signals of need increased participants' belief in the victim and their likelihood of helping her, with the increase in perceived need and likelihood of helping increasing monotonically with signal cost. As predicted, the increase in likelihood of helping was largely mediated by the increase in belief in the victim. In an exploratory analysis, costlier signals also decreased perceptions that the victim was manipulative. These results provide evidence that, contrary to the influential "interpersonal" view that depressive behaviors are socially dysfunctional [reviewed in @hames2013], they in fact outperform verbal requests, sad expressions, and crying in providing benefits to victims when there are conflicts of interest.

Signal effects were largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters, and smaller in the "basketball coach" and the "thwarted marriage" vignettes. The small effect in the "basketball coach" vignette might have been due to the lack of relatedness between participants and the victim (the star player), or how participants weighted the costs of suspending the coach vs. punishing a potentially innocent person. The US vignettes also had different degrees of evidence *against* the victim beyond just denial by the accused, ranging from strong evidence in the basketball coach vignette (a negative police report) to moderate evidence in the romantic partner vignette (no physical injuries) to weak evidence in the brother-in-law vignette (nothing beyond denial by the brother-in-law), raising the possibility that credible signals are more effective when negative evidence is lacking (Dylan Tweed, personal communication).

The small signal effect in the "thwarted marriage" vignette, which involved the Indian sample, could indicate that our results do not generalize across cultures, undermining our adaptationist hypothesis. It could also reflect our poor understanding of contemporary Indian culture regarding dowry (the effect was larger in older participants). Baseline belief in the older daughter, and likelihood of helping her, was relatively high at baseline (`r signif(100*mean(d$T1Belief[d$vignette=='Thwarted marriage']), 2)`%) compared to victims in the other vignettes. Private information and conflict therefore probably played a smaller role and thus costly signals were less necessary. We observed a similar pattern in our pilot study, in which baseline belief in the victim's need was high, and costly signals had smaller effects than they did in the current study. Additionally, supporting the older daughter came at the cost of one's younger daughter, which may also help explain the relatively small signal effects. A final consideration is that data from the Indian sample appeared to be of lower quality, limiting our confidence in any of these interpretations (see the Limitations section for more information).

In the "brother-in-law" and "romantic partner" scenarios, *Crying* had little effect on the magnitude of pro-victim responses relative to *Verbal request*, suggesting it was not costly enough to serve as a reliable signal in times of substantial conflicts of interests. In contrast, both *Depression* conditions increased support, albeit to similar degrees. One potential reason for the similar effects of the *Depression* conditions is the increase in costs from *Mild depression* to *Depression* was small (e.g., grades dropping from As to Bs in *Mild depression* vs. Cs in *Depression*). Such small changes may be less impactful in vignettes than in real-life, where the effects of signaling may increase in severity as they persist over time.

@@

Contrary to our adaptationist hypothesis, and supporting the mainstream view that depression is a psychopathology, participants' perceptions that the victim was mentally ill increased with signal cost. However, there have been extensive media campaigns to convince the public that depression is a mental illness with the laudable goal of reducing stigma [@corrigan2012; @rusch2005]. Even so, in the *Depression* conditions across vignettes, no more than 25% of participants thought the victim was mentally ill, and in the *Suicide attempt* condition the proportion of participants perceiving mental illness exceeded 50% only in the Basketball coach vignette. Although perceived mental illness was associated with somewhat lower *T2 Belief* and *T2 Action*, this effect was mainly evident in the *Verbal request* and *Crying* conditions.

@@

Depression is costly and sometimes leads to death by suicide. Our results suggest that these costs, which mainstream theories take as evidence of brain dysfunction, instead function to help victims of adversity elicit support when their true level of need is private information and they have conflicts with social partners. These findings align with real-world evidence that depression and suicide simultaneously elicit positive and negative responses from social partners [for review, see @SymeHagen2021]. If the bargaining model is correct, depression and suicidality are important social problems, not medical ones [@Hagen2003; @syme2020].

@@

In the MTurk pilot study, participants read a vignette that put them in situations in which an imagined sister was requesting a loan of \$50,000 from the participant that was originally saved for the participant's imagined daughter's college education. The sister claims the money is needed for medical treatment for her own child, but she has incentives to deceive. The manipulations were signals very similar to those used in this study, and the outcomes were belief that the sister really needed the money and intention to loan her the money. See XX for the full write-up. See <https://osf.io/3rg8b> for preregistration of the pilot study. See <http://doi.org/10.5281/zenodo.4637883> for pilot data.

@@

```{r mentallyillT2, fig.width=8, fig.cap="Proportion of participants in each condition who thought that the victim was mentally ill at T2, controlling for perceptions of mental illness at T1. Means estimated with a generalized linear regression model with the quasibinomial family, and displayed for participants who did not perceive mental illness at T1."}
effectsPlots$m21plot + ggtitle('Time 2')
```

```{r mentallyillbelief, fig.width=8, fig.cap="Association of T2 Perceived Mental Illness (Yes/No) with T2 Belief in the US sample, by signal, controlling for T1 Perceived Mental Illness (Yes/No), T1 Belief, and vignette. Means estimated with a generalized linear regression model with the quasibinomial family."}
effectsPlots_US$m28plot +
  xlab('\nPerceived mental illness') +
  theme(axis.text.x = element_text(angle = 0, hjust=0.5))
```

```{r mentallyillaction, fig.width=8, fig.cap="Association of T2 Perceived Mental Illness (Yes/No) with T2 Action in the US sample, by signal, controlling for T1 Perceived Mental Illness (Yes/No), T1 Action, and vignette. Means estimated with a generalized linear regression model with the quasibinomial family."}
effectsPlots_US$m29plot +
  xlab('\nPerceived mental illness') +
  theme(axis.text.x = element_text(angle = 0, hjust=0.5))
```
